---
title: "R Notebook"
output: html_notebook
---

### Load libraries
```{r}
library(DESeq2)
library(vsn)
library(ggplot2)
library(tidyr)
library(dplyr)
```

# Tolerance only analysis

#### Load counts data and meta-data
```{r}
Counts1 <- read.table(file="Comp1readCountTDM_MERS_github.txt", sep="\t", header=T)

names(Counts1) <- gsub("Aligned.sortedByCoord.out.bam", "", names(Counts1))

StudyDes1 <- read.table(file="StudyDesignComp1_github.txt", sep="\t", header = TRUE)
StudyDes1 <- StudyDes1[, -4] # remove column with sex of samples
```

#### Convert to matrix, filter out low read genes
```{r}
ddsMat1 <- DESeqDataSetFromMatrix(countData = Counts1, colData = StudyDes1, design = ~Collection_Date)
nrow(ddsMat1)
keep <- rowSums(counts(ddsMat1) >= 15) >= 6 # keeps genes that have 15 or more counts in at least 6 samples
ddsMat1 <- ddsMat1[keep,]
nrow(ddsMat1)
```

## Part I. MDS plot of normalized counts for all genes

#### Variance-stabalize all genes, calculate Euclidean distances
```{r}
# apply variance-stabilizing transformation, blind = FALSE to preserve meta-data grouping
vsd1 <- vst(ddsMat1, blind = FALSE)
head(assay(vsd1))

# calculate Euclidean distances
sampleDists1 <- dist(t(assay(vsd1)))

# convert to matrix
sampleDistMatrix1 <- as.matrix(sampleDists1)
rownames(sampleDistMatrix1) <- paste(StudyDes1$Collection_Date, sep="-")

# multidimensional scaling (MDS) of data matrix
mds1 <- data.frame(cmdscale(sampleDistMatrix1))
mds1 <- cbind(mds1, colData(vsd1))
```

#### Generate MDS plot for normalized genes
```{r}
# split 'Collection_Date' into Day and Condition columns
mds1_filtered <- mds1 %>%
  separate(Collection_Date, into = c("Day", "Condition"), sep = "_") %>%
  filter(!is.na(Condition))

# define color based on Condition
condition_colours <- c("HIGH" = "red", "LOW" = "blue", "MOCK" = "darkgrey")

# define shape based on Day
day_shapes <- c(
  "Day1" = 1,  # open circle
  "Day2" = 2,  # open triangle
  "Day3" = 15,  # square
  "Day4" = 17,  # triangle
  "Day5" = 16)   # circle 

# generate the MDS Plot
ggplot(mds1_filtered, aes(x = X1, y = X2, shape = Day, color = Condition)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = condition_colours,
                     labels = c("High dose", "Low dose", "Mock")) +
  scale_shape_manual(values = day_shapes,
                     labels = c("1", "2", "3", "4", "5")) +
  theme_classic() +
  labs(x = "Dimension 1", 
       y = "Dimension 2") +
  theme(text = element_text(family = "Times"),
        legend.position = "right",
        legend.title = element_text(size = 17, face = "bold"),
        legend.text = element_text(size = 16),
        axis.title = element_text(size = 20, face = "bold"),
        axis.text = element_text(size = 17),
        axis.line = element_line(size = 0.1, color = "black"),
        panel.border = element_rect(color = "black", fill = NA, size = 0.5),
        aspect.ratio = 0.75)

ggsave("MDS_Plot_Tolerance_normalized.png", width = 8, height = 6, dpi = 300)
```
## Part II. MDS plot of normalized counts for differentially expressed (DE) genes

#### Run DESeq2 analysis
```{r}
# 0.05 p-value cut-off
DEMat1 <- DESeq(ddsMat1)
res1 <- results(DEMat1, alpha = 0.05) 
summary(res1)

# define DE gene expression as p < 0.05 and Log2FC > |1.5|
DE_genes1 <- rownames(res1[!is.na(res1$padj) & !is.na(res1$log2FoldChange) &
                           res1$padj < 0.05 & abs(res1$log2FoldChange) > 1.5, ])
```

#### Variance-stabalize DE genes, calculate Euclidean distances
```{r}
# apply variance-stabilizing transformation, blind = FALSE to preserve meta-data grouping
vsd1_de <- vst(DEMat1, blind = FALSE) 
vsd1_de_cut_off <- assay(vsd1)[DE_genes1, ] # subset VST data to include DE genes only
head(vsd1_de_cut_off)

# calculate Euclidean distances 
sampleDists_de1 <- dist(t(vsd1_de_cut_off))

# convert to matrix
sampleDistMatrix_de1 <- as.matrix(sampleDists_de1)
rownames(sampleDistMatrix_de1) <- paste(StudyDes1$Collection_Date, sep="-")

# multidimensional scaling (MDS) of data matrix
mds_de1 <- data.frame(cmdscale(sampleDistMatrix_de1))
mds1_de1 <- cbind(mds_de1, colData(vsd1_de))
```

#### Generate MDS plot for normalized DE genes
```{r}
# split 'Collection_Date' into Day and Condition columns
mds_data_de1_filtered <- mds1_de1 %>%
  separate(Collection_Date, into = c("Day", "Condition"), sep = "_") %>%
  filter(Condition != "MOCK") 

# define color based on Condition
condition_colors1 <- c("HIGH" = "red", "LOW" = "blue")

# define shape based on Day
day_shapes <- c(
  "Day1" = 1,  # open circle
  "Day2" = 2,  # open triangle
  "Day3" = 15,  # square
  "Day4" = 17,  # triangle
  "Day5" = 16)   # circle

# generate the MDS plot
ggplot(mds_data_de1_filtered, aes(x = X1, y = X2, color = Condition, shape = Day)) +
  geom_point(size = 4, stroke = 2) +
  scale_color_manual(values = condition_colors1,
                     labels = c("High dose", "Low dose")) +
  scale_shape_manual(values = day_shapes,
                     labels = c("1", "2", "3", "4", "5")) +
  theme_classic() +
  labs(x = "Dimension 1", 
       y = "Dimension 2") +
  theme(text = element_text(family = "Times"),
        legend.position = "bottom",
        legend.box = "vertical",
        legend.title = element_text(size = 18, face = "bold"),
        legend.text = element_text(size = 17),
        axis.title = element_text(size = 20, face = "bold"),
        axis.text = element_text(size = 17),
        axis.line = element_line(size = 0.1, color = "black"),
        panel.border = element_rect(color = "black", fill = NA, size = 0.5))

ggsave("MDS_Plot_Tolerance_DE_1.png", width = 8, height = 7, dpi = 300)
```

# Tolerance and sex analysis

#### Load counts data and meta-data
```{r}
Counts2 <- read.table(file="Comp2readCountTDM_MERS_github.txt", sep="\t", header = TRUE, check.names = FALSE)

names(Counts2) <- gsub("Aligned.sortedByCoord.out.bam", "", names(Counts2))

StudyDes2 <- read.table(file="StudyDesignComp2_github.txt", sep="\t", header = TRUE)
```

#### Convert to matrix, filter out low read genes
```{r}
ddsMat2 <- DESeqDataSetFromMatrix(countData = Counts2, colData = StudyDes2, design = ~Collection_Date)
nrow(ddsMat2)
keep <- rowSums(counts(ddsMat2) >= 15) >= 6 # keeps genes that have 15 or more counts in at least 6 samples
ddsMat <- ddsMat2[keep,]
nrow(ddsMat2)
```
## Part I. MDS plot of normalized counts for all genes

#### Variance-stabalize all genes, calculate Euclidean distances
```{r}
# apply variance-stabilizing transformation, blind = FALSE to preserve meta-data grouping
vsd2 <- vst(ddsMat2, blind = FALSE)
head(assay(vsd2))

# calculate Euclidean distances
sampleDists2 <- dist(t(assay(vsd2)))

# convert to matrix
sampleDistMatrix2 <- as.matrix(sampleDists2)
rownames(sampleDistMatrix2) <- paste(StudyDes2$Collection_Date, sep="-")

# multidimensional scaling (MDS) of data matrix
mds2 <- data.frame(cmdscale(sampleDistMatrix2))
mds2 <- cbind(mds2, colData(vsd2))
```

#### Generate MDS plot for normalized genes
```{r}
# split 'Collection_Date' into Day, Condition and Sex columns
mds2_filtered <- mds2 %>%
  separate(Collection_Date, into = c("Day", "Condition", "Sex"), sep = "_") %>%
  filter(!is.na(Condition))

# merge Condition and Sex again
mds2_filtered$Condition_Sex <- paste(mds2_filtered$Condition, mds2_filtered$Sex, sep = "_")

# define color based on Condition and Sex
condition_sex_colors <- c("HIGH_F" = "red", "LOW_F" = "blue",  "HIGH_M" = "sienna1",  "LOW_M" = "cadetblue3",  
                          "MOCK_F" = "gray61","MOCK_M" = "black")

# define shape based on Day
day_shapes <- c(
  "Day1" = 1,  # open circle
  "Day2" = 2,  # open triangle
  "Day3" = 15,  # square
  "Day4" = 17,  # triangle
  "Day5" = 16)   # circle

# generate the MDS Plot
ggplot(mds2_filtered, aes(x = X1, y = X2, shape = Day, color = Condition_Sex)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_manual(values = condition_sex_colors,
                     labels = c("High dose female", "High dose male", "Low dose female", "Low dose male", "Mock female", "Mock male", "Day 1")) +
  scale_shape_manual(values = day_shapes,
                     labels = c("1", "2", "3", "4", "5")) +
  theme_classic() +
  labs(x = "Dimension 1", 
       y = "Dimension 2") +
  theme(text = element_text(family = "Times"),
        legend.position = "right",
        legend.title = element_text(size = 17, face = "bold"),
        legend.text = element_text(size = 16),
        axis.title = element_text(size = 20, face = "bold"),
        axis.text = element_text(size = 15),
        axis.line = element_line(size = 0.1, color = "black"),
        panel.border = element_rect(color = "black", fill = NA, size = 0.5),
        aspect.ratio = 0.75) +
  guides(color = guide_legend(title = "Condition", order = 1),  # Change legend title for color
         shape = guide_legend(title = "Day", order = 2)) +
  scale_fill_discrete(labels=c('High dose female', 'High dose male', 'Low dose female', 'Low dose male')) 

ggsave("MDS_Plot_Tolerance&Sex_normalized_2.png", width = 8, height = 6, dpi = 300)
```
## Part II. MDS plot of normalized counts for differentially expressed (DE) genes

#### Run DESeq2 analysis
```{r}
# 0.05 p-value cut-off
DEMat2 <- DESeq(ddsMat2)
res2 <- results(DEMat2, alpha = 0.05) # p-value cut-off
summary(res2)

# define DE gene expression as p < 0.05 and Log2FC > |1.5|
DE_genes2 <- rownames(res2[!is.na(res2$padj) & !is.na(res2$log2FoldChange) &
                           res2$padj < 0.05 & abs(res2$log2FoldChange) > 1.5, ])
```


# Variance-stabalize DE genes, calculate Euclidean distances
```{r}
# apply variance stabilizing transformation (VST)
vsd2_de <- vst(DEMat2, blind = FALSE) 
vsd2_de_cut_off <- assay(vsd2_de)[DE_genes2, ] # subset VST data to include DE genes only
head(vsd2_de_cut_off)

# calculate Euclidean distances 
sampleDists_de2 <- dist(t(vsd2_de_cut_off))

# convert to matrix
sampleDistMatrix_de2 <- as.matrix(sampleDists_de2)
rownames(sampleDistMatrix_de2) <- paste(StudyDes2$Collection_Date, sep="-")

# multidimensional scaling (MDS) of data matrix
mds_de2 <- cmdscale(sampleDists_de2, k = 2)

mds2_de2 <- data.frame(
  Dim1 = mds_de2[,1],
  Dim2 = mds_de2[,2],
  as.data.frame(colData(vsd2_de))
)
```

#### Generate MDS plot for normalized DE genes
```{r}
# split 'Collection_Date' into Day, Condition, and Sex columns
mds_data_de2_filtered <- mds2_de2 %>%
  separate(Collection_Date, into = c("Day", "Condition", "Sex"), sep = "_") %>%
  filter(Condition != "MOCK")

# merge Condition and Sex again
mds_data_de2_filtered$Condition_Sex <- paste(mds_data_de2_filtered$Condition, mds_data_de2_filtered$Sex, sep = "_")

# define color based on Condition and Sex
condition_sex_colors <- c("HIGH_F" = "red",  "LOW_F" = "blue", "HIGH_M" = "sienna1", "LOW_M" = "cadetblue3")

# define shape based on Day
day_shapes <- c(
  "Day1" = 1,  # open circle
  "Day2" = 2,  # open triangle
  "Day3" = 15,  # square
  "Day4" = 17,  # triangle
  "Day5" = 16)   # circle

# Generate the MDS Plot
ggplot(mds_data_de2_filtered, aes(x = Dim1, y = Dim2, shape = Day, color = Condition_Sex)) +
  geom_point(size = 4, stroke = 2) +
  scale_color_manual(values = condition_sex_colors,
                     labels = c("High dose female", "High dose male", "Low dose female", "Low dose male", "Day 1")) +
  scale_shape_manual(values = day_shapes,
                     labels = c("1", "2", "3", "4", "5")) +
  theme_classic() +
  labs(x = "Dimension 1", 
       y = "Dimension 2") +
  theme(text = element_text(family = "Times"),
        legend.position = "bottom",
        legend.box = "vertical",
        legend.title = element_text(size = 17, face = "bold"),
        legend.text = element_text(size = 16),
        axis.title = element_text(size = 20, face = "bold"),
        axis.text = element_text(size = 16),
        axis.line = element_line(size = 0.1, color = "black"),
        panel.border = element_rect(color = "black", fill = NA, size = 0.5),
        aspect.ratio = 0.8) +
  guides(color = guide_legend(title = "Condition"),
         shape = guide_legend(title = "Day")) +
  scale_fill_discrete(labels=c('High dose female', 'High dose male', 'Low dose female', 'Low dose male')) 

ggsave("MDS_Plot_Tolerance&Sex_DE.png", width = 10, height = 7, dpi = 300)
```

