---
title: "Volcano"
author: "Marin H"
date: "2026-02-13"
output: html_document
---

## load libraries
```{r}
library(ggplot2)
library(dplyr)
library(DESeq2)
library(ggtext)
```

# Tolerance only analysis

#### Load counts data and meta-data
```{r}
Counts1 <- read.table(file="/Users/mh_mac/Documents/Usask - MSc/Graduate/Research/Computational/MERS paper/GitHub/MDS/Comp1readCountTDM_MERS_github.txt", sep="\t", header=T)

names(Counts1) <- gsub("Aligned.sortedByCoord.out.bam", "", names(Counts1))

StudyDes1 <- read.table(file="/Users/mh_mac/Documents/Usask - MSc/Graduate/Research/Computational/MERS paper/GitHub/MDS/StudyDesignComp1_github.txt", sep="\t", header = TRUE)
StudyDes1 <- StudyDes1[, -4] # remove column with sex of samples
```

#### Convert to matrix, filter out low read genes
```{r}
ddsMat1 <- DESeqDataSetFromMatrix(countData = Counts1, colData = StudyDes1, design = ~Collection_Date)
nrow(ddsMat1)
keep <- rowSums(counts(ddsMat1) >= 15) >= 6 # keeps genes that have 15 or more counts in at least 6 samples
ddsMat1 <- ddsMat1[keep,]
nrow(ddsMat1)
```

#### Run DE analysis
```{r}
DE_1 <- DESeq(ddsMat1)
```

## Part I. Generate individual and combined volcano plots

#### Extract results for specific comparisons
```{r}
# Low vs. Mock
Day1_LOWvsMOCK <- results(DE_1, alpha = 0.05, contrast = c("Collection_Date", "Day1_LOW", "Day1_MOCK"))
Day1_LOWvsMOCK_df <- as.data.frame(Day1_LOWvsMOCK)
Day1_LOWvsMOCK_df$Gene <- rownames(Day1_LOWvsMOCK_df)
Day1_LOWvsMOCK_df$Day <- "Day1"
Day1_LOWvsMOCK_df$Condition <- "Low"

Day2_LOWvsMOCK <- results(DE_1, alpha = 0.05, contrast = c("Collection_Date", "Day2_LOW", "Day2_MOCK"))
Day2_LOWvsMOCK_df <- as.data.frame(Day2_LOWvsMOCK)
Day2_LOWvsMOCK_df$Gene <- rownames(Day2_LOWvsMOCK_df)
Day2_LOWvsMOCK_df$Day <- "Day2"
Day2_LOWvsMOCK_df$Condition <- "Low"

Day3_LOWvsMOCK <- results(DE_1, alpha = 0.05, contrast = c("Collection_Date", "Day3_LOW", "Day3_MOCK"))
Day3_LOWvsMOCK_df <- as.data.frame(Day3_LOWvsMOCK)
Day3_LOWvsMOCK_df$Gene <- rownames(Day3_LOWvsMOCK_df)
Day3_LOWvsMOCK_df$Day <- "Day3"
Day3_LOWvsMOCK_df$Condition <- "Low"

Day4_LOWvsMOCK <- results(DE_1, alpha = 0.05, contrast = c("Collection_Date", "Day4_LOW", "Day4_MOCK"))
Day4_LOWvsMOCK_df <- as.data.frame(Day4_LOWvsMOCK)
Day4_LOWvsMOCK_df$Gene <- rownames(Day4_LOWvsMOCK_df)
Day4_LOWvsMOCK_df$Day <- "Day4"
Day4_LOWvsMOCK_df$Condition <- "Low"

Day5_LOWvsMOCK <- results(DE_1, alpha = 0.05, contrast = c("Collection_Date", "Day5_LOW", "Day5_MOCK"))
Day5_LOWvsMOCK_df <- as.data.frame(Day5_LOWvsMOCK)
Day5_LOWvsMOCK_df$Gene <- rownames(Day5_LOWvsMOCK_df)
Day5_LOWvsMOCK_df$Day <- "Day5"
Day5_LOWvsMOCK_df$Condition <- "Low"


# High vs. Mock
Day1_HIGHvsMOCK <- results(DE_1, alpha = 0.05, contrast = c("Collection_Date", "Day1_HIGH", "Day1_MOCK"))
Day1_HIGHvsMOCK_df <- as.data.frame(Day1_HIGHvsMOCK)
Day1_HIGHvsMOCK_df$Gene <- rownames(Day1_HIGHvsMOCK_df)
Day1_HIGHvsMOCK_df$Day <- "Day1"
Day1_HIGHvsMOCK_df$Condition <- "High"


Day2_HIGHvsMOCK <- results(DE_1, alpha = 0.05, contrast = c("Collection_Date", "Day2_HIGH", "Day2_MOCK"))
Day2_HIGHvsMOCK_df <- as.data.frame(Day2_HIGHvsMOCK)
Day2_HIGHvsMOCK_df$Gene <- rownames(Day2_HIGHvsMOCK_df)
Day2_HIGHvsMOCK_df$Day <- "Day2"
Day2_HIGHvsMOCK_df$Condition <- "High"


Day3_HIGHvsMOCK <- results(DE_1, alpha = 0.05, contrast = c("Collection_Date", "Day3_HIGH", "Day3_MOCK"))
Day3_HIGHvsMOCK_df <- as.data.frame(Day3_HIGHvsMOCK)
Day3_HIGHvsMOCK_df$Gene <- rownames(Day3_HIGHvsMOCK_df)
Day3_HIGHvsMOCK_df$Day <- "Day3"
Day3_HIGHvsMOCK_df$Condition <- "High"


Day4_HIGHvsMOCK <- results(DE_1, alpha = 0.05, contrast = c("Collection_Date", "Day4_HIGH", "Day4_MOCK"))
Day4_HIGHvsMOCK_df <- as.data.frame(Day4_HIGHvsMOCK)
Day4_HIGHvsMOCK_df$Gene <- rownames(Day4_HIGHvsMOCK_df)
Day4_HIGHvsMOCK_df$Day <- "Day4"
Day4_HIGHvsMOCK_df$Condition <- "High"


Day5_HIGHvsMOCK <- results(DE_1, alpha = 0.05, contrast = c("Collection_Date", "Day5_HIGH", "Day5_MOCK"))
Day5_HIGHvsMOCK_df <- as.data.frame(Day5_HIGHvsMOCK)
Day5_HIGHvsMOCK_df$Gene <- rownames(Day5_HIGHvsMOCK_df)
Day5_HIGHvsMOCK_df$Day <- "Day5"
Day5_HIGHvsMOCK_df$Condition <- "High"


# combine results from all comparisons
DE1_results_LOW_df <- rbind(Day1_LOWvsMOCK_df, Day2_LOWvsMOCK_df, Day3_LOWvsMOCK_df,
                            Day4_LOWvsMOCK_df, Day5_LOWvsMOCK_df)
DE1_results_HIGH_df <- rbind(Day1_HIGHvsMOCK_df, Day2_HIGHvsMOCK_df, Day3_HIGHvsMOCK_df,
                             Day4_HIGHvsMOCK_df, Day5_HIGHvsMOCK_df)

DE1_results_ALL_df <- rbind(Day1_LOWvsMOCK_df, Day2_LOWvsMOCK_df, Day3_LOWvsMOCK_df, 
                            Day4_LOWvsMOCK_df, Day5_LOWvsMOCK_df, Day1_HIGHvsMOCK_df,
                            Day2_HIGHvsMOCK_df, Day3_HIGHvsMOCK_df, Day4_HIGHvsMOCK_df,
                            Day5_HIGHvsMOCK_df)
```

#### Generate plotting function
```{r}
plot_volcano <- function(DE1_results_ALL_df, day_label, condition_label, colour_label) {
  
  # add 'Day' and 'Condition' columns
  DE1_results_ALL_df$Day <- day_label
  DE1_results_ALL_df$Condition <- condition_label
  
  # create the volcano plot
  volcano_plot <- ggplot(DE1_results_ALL_df, aes(x = log2FoldChange, y = -log10(pvalue))) +
    geom_point(aes(color = (padj < 0.05  & abs(log2FoldChange) > 1.5)), alpha = 0.6, size = 2) +
    scale_color_manual(values = c("gray", colour_label)) +
    theme_classic() +
    ggtitle(paste(day_label, ":", condition_label, "vs Mock")) +
    xlab(expression(Log[2]~"(Fold Change)")) +
    ylab(expression(-Log[10]~"("*italic(P)*"-value)")) +
    theme(legend.position = "none",
          text = element_text(family = "Times"))
  
  return(volcano_plot)
}
```

#### Plot individual volcano plots
```{r}
# Low vs. Mock
ggsave("Volcano_indiv_Day1_LvM.png", plot_volcano(Day1_LOWvsMOCK_df, "Day1", "LOW", "blue"), width = 5, height = 3.5, dpi = 300)
ggsave("Volcano_indiv_Day2_LvM.png", plot_volcano(Day2_LOWvsMOCK_df, "Day2", "LOW", "blue"), width = 5, height = 3.5, dpi = 300)
ggsave("Volcano_indiv_Day3_LvM.png", plot_volcano(Day3_LOWvsMOCK_df, "Day3", "LOW", "blue"), width = 5, height = 3.5, dpi = 300)
ggsave("Volcano_indiv_Day4_LvM.png", plot_volcano(Day4_LOWvsMOCK_df, "Day4", "LOW", "blue"), width = 5, height = 3.5, dpi = 300)
ggsave("Volcano_indiv_Day5_LvM.png", plot_volcano(Day5_LOWvsMOCK_df, "Day5", "LOW", "blue"), width = 5, height = 3.5, dpi = 300)

# High vs. Mock
ggsave("Volcano_indiv_Day1_HvM.png", plot_volcano(Day1_HIGHvsMOCK_df, "Day1", "HIGH", "red"), width = 5, height = 3.5, dpi = 300)
ggsave("Volcano_indiv_Day2_HvM.png", plot_volcano(Day2_HIGHvsMOCK_df, "Day2", "HIGH", "red"), width = 5, height = 3.5, dpi = 300)
ggsave("Volcano_indiv_Day3_HvM.png", plot_volcano(Day3_HIGHvsMOCK_df, "Day3", "HIGH", "red"), width = 5, height = 3.5, dpi = 300)
ggsave("Volcano_indiv_Day4_HvM.png", plot_volcano(Day4_HIGHvsMOCK_df, "Day4", "HIGH", "red"), width = 5, height = 3.5, dpi = 300)
ggsave("Volcano_indiv_Day5_HvM.png", plot_volcano(Day5_HIGHvsMOCK_df, "Day5", "HIGH", "red"), width = 5, height = 3.5, dpi = 300)
```

#### Generate combined volcano plot
```{r}
# p-value < 0.05 and Log2FC > |1.5|
DE1_results_ALL_df_plot_1 <- DE1_results_ALL_df %>%
  mutate(sigGroup = ifelse(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1.5,
                           as.character(Condition),
                           "Not significant")) %>%
  mutate(Day = gsub("Day", "Day ", Day))
table(DE1_results_ALL_df_plot_1$sigGroup)

# plot
combined_volcano_plot <- ggplot(DE1_results_ALL_df_plot_1, 
                                aes(x = log2FoldChange, y = -log10(pvalue), 
                                    color = sigGroup)) +
  geom_point(alpha = 0.2, size = 2) +
  scale_color_manual(
    name = "Condition",
    values = c("High" = "red", "Low" = "blue", "NotSig" = "gray"),
    labels = c("High dose", "Low dose")) +
  facet_wrap(~Day, scales = "free_x") +
  theme_classic() +
  ggtitle("Combined Volcano Plot: Low vs MOCK and High vs MOCK") +
  xlab(expression(bold(Log[2]~"(Fold Change)"))) +
  ylab(expression(bold(-Log[10]~"("*italic(p)*"-value)"))) +
  theme(text = element_text(family = "Times"),
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 13),
    legend.title = element_blank(),
    legend.text = element_text(size = 13),
    legend.position = "bottom",
    strip.text = element_markdown(size = 14, face = "bold"),  
    panel.border = element_rect(color = "black", fill = NA, size = 0.4),
    axis.line = element_line(size = 0.1, color = "black"),
    strip.background = element_rect(color = "black", fill = "white", size = 0.4))

ggsave("Volcano_Tolerance.png", combined_volcano_plot, width = 10, height = 7, dpi = 300)

print(combined_volcano_plot)
```

## Part II. Generate bar-coded volcano plot

#### Sort DE genes to up- and down-regulated, determine counts
```{r}
# sort DE genes
DE1_results_ALL_df_plot_2 <- DE1_results_ALL_df %>%
  mutate(Regulation = case_when(
    !is.na(padj) & padj < 0.05 & log2FoldChange > 1.5 ~ "Upregulated",
    !is.na(padj) & padj < 0.05 & log2FoldChange < -1.5 ~ "Downregulated",
    TRUE ~ "Not Significant"
  )) %>%
  mutate(Day = gsub("Day", "", Day),
         Group = case_when(
           Condition == "High" ~ "High dose",
           Condition == "Low" ~ "Low dose"))

# count the number of up- and down-regulated genes per condition and day
bar_data_1 <- DE1_results_ALL_df_plot_2 %>%
  filter(Regulation != "Not Significant") %>%
  group_by(Day, Condition, Regulation) %>%
  summarise(Gene_Count = n(), .groups = "drop")
```

#### Plot
```{r}
# add labels and colour by condition
bar_data_1 <- bar_data_1 %>%
  mutate(Group_colored = case_when(
    Condition == "High" ~ "<span style='color:red;'>High dose</span>",
    Condition == "Low" ~ "<span style='color:blue;'>Low dose</span>"
  )) %>%
  mutate(Group_colored = factor(Group_colored, levels = c(
    "<span style='color:red;'>High dose</span>",
    "<span style='color:blue;'>Low dose</span>"
  )))

# plot
bar_plot_1 <- ggplot(bar_data_1, aes(x = Day, y = Gene_Count, fill = Regulation)) +
  geom_bar(stat = "identity", position = "stack", width = 0.90) +
  facet_wrap(~ Group_colored) +  
  scale_fill_manual(values = c("Upregulated" = "orange", "Downregulated" = "purple")) +
  labs(title = "Differentially Expressed Genes Over Time",
       x = "Day",
       y = "Gene Count",
       fill = "Regulation") +
  theme_classic() +
  theme(
    text = element_text(family = "Times"),
    axis.title = element_text(size = 17, face = "bold"),
    axis.text = element_text(size = 13),
    legend.title = element_text(size = 15, face = "bold"),
    legend.text = element_text(size = 13),
    legend.position = "bottom",
    strip.text = element_markdown(size = 14, face = "bold"),  
    panel.border = element_rect(color = "black", fill = NA, size = 0.4),
    axis.line = element_line(size = 0.1, color = "black"),
    strip.background = element_rect(color = "black", fill = "white", size = 0.4))  

ggsave("Bar_Graph_Tolerance_2.0.png", bar_plot_1, width = 6, height = 4, dpi = 300)
print(bar_data_1)
print(bar_plot_1)
```


# Tolerance and sex analysis

#### Load counts data and meta-data
```{r}
Counts2 <- read.table(file="/Users/mh_mac/Documents/Usask - MSc/Graduate/Research/Computational/MERS paper/GitHub/MDS/Comp2readCountTDM_MERS_github.txt", sep="\t", header = TRUE, check.names = FALSE)

names(Counts2) <- gsub("Aligned.sortedByCoord.out.bam", "", names(Counts2))

StudyDes2 <- read.table(file="/Users/mh_mac/Documents/Usask - MSc/Graduate/Research/Computational/MERS paper/GitHub/MDS/StudyDesignComp2_github.txt", sep="\t", header = TRUE)
```

#### Convert to matrix, filter out low read genes
```{r}
ddsMat2 <- DESeqDataSetFromMatrix(countData = Counts2, colData = StudyDes2, design = ~Collection_Date)
nrow(ddsMat2)
keep <- rowSums(counts(ddsMat2) >= 15) >= 6 # keeps genes that have 15 or more counts in at least 6 samples
ddsMat <- ddsMat2[keep,]
nrow(ddsMat2)
```
#### Run DE analysis
```{r}
DE_2 <- DESeq(ddsMat2)
```

## Part I. Generate individual and combined volcano plots

#### Extract results for specific comparisons
```{r}
# Low vs. Mock Female
Day1_LOWvsMOCK_F <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day1_LOW_F", "Day1_MOCK_F"))
Day1_LOWvsMOCK_F_df <- as.data.frame(Day1_LOWvsMOCK_F)
Day1_LOWvsMOCK_F_df$Gene <- rownames(Day1_LOWvsMOCK_F_df)
Day1_LOWvsMOCK_F_df$Day <- "Day1"
Day1_LOWvsMOCK_F_df$Condition <- "Low"
Day1_LOWvsMOCK_F_df$Sex <- "Female"

Day2_LOWvsMOCK_F <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day2_LOW_F", "Day2_MOCK_F"))
Day2_LOWvsMOCK_F_df <- as.data.frame(Day2_LOWvsMOCK_F)
Day2_LOWvsMOCK_F_df$Gene <- rownames(Day2_LOWvsMOCK_F_df)
Day2_LOWvsMOCK_F_df$Day <- "Day2"
Day2_LOWvsMOCK_F_df$Condition <- "Low"
Day2_LOWvsMOCK_F_df$Sex <- "Female"

Day3_LOWvsMOCK_F <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day3_LOW_F", "Day3_MOCK_F"))
Day3_LOWvsMOCK_F_df <- as.data.frame(Day3_LOWvsMOCK_F)
Day3_LOWvsMOCK_F_df$Gene <- rownames(Day3_LOWvsMOCK_F_df)
Day3_LOWvsMOCK_F_df$Day <- "Day3"
Day3_LOWvsMOCK_F_df$Condition <- "Low"
Day3_LOWvsMOCK_F_df$Sex <- "Female"

Day4_LOWvsMOCK_F <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day4_LOW_F", "Day4_MOCK_F"))
Day4_LOWvsMOCK_F_df <- as.data.frame(Day4_LOWvsMOCK_F)
Day4_LOWvsMOCK_F_df$Gene <- rownames(Day4_LOWvsMOCK_F_df)
Day4_LOWvsMOCK_F_df$Day <- "Day4"
Day4_LOWvsMOCK_F_df$Condition <- "Low"
Day4_LOWvsMOCK_F_df$Sex <- "Female"

Day5_LOWvsMOCK_F <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day5_LOW_F", "Day5_MOCK_F"))
Day5_LOWvsMOCK_F_df <- as.data.frame(Day5_LOWvsMOCK_F)
Day5_LOWvsMOCK_F_df$Gene <- rownames(Day5_LOWvsMOCK_F_df)
Day5_LOWvsMOCK_F_df$Day <- "Day5"
Day5_LOWvsMOCK_F_df$Condition <- "Low"
Day5_LOWvsMOCK_F_df$Sex <- "Female"

# Low vs. Mock Male
Day1_LOWvsMOCK_M <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day1_LOW_M", "Day1_MOCK_M"))
Day1_LOWvsMOCK_M_df <- as.data.frame(Day1_LOWvsMOCK_M)
Day1_LOWvsMOCK_M_df$Gene <- rownames(Day1_LOWvsMOCK_M_df)
Day1_LOWvsMOCK_M_df$Day <- "Day1"
Day1_LOWvsMOCK_M_df$Condition <- "Low"
Day1_LOWvsMOCK_M_df$Sex <- "Male"

Day2_LOWvsMOCK_M <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day2_LOW_M", "Day2_MOCK_M"))
Day2_LOWvsMOCK_M_df <- as.data.frame(Day2_LOWvsMOCK_M)
Day2_LOWvsMOCK_M_df$Gene <- rownames(Day2_LOWvsMOCK_M_df)
Day2_LOWvsMOCK_M_df$Day <- "Day2"
Day2_LOWvsMOCK_M_df$Condition <- "Low"
Day2_LOWvsMOCK_M_df$Sex <- "Male"

Day3_LOWvsMOCK_M <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day3_LOW_M", "Day3_MOCK_M"))
Day3_LOWvsMOCK_M_df <- as.data.frame(Day3_LOWvsMOCK_M)
Day3_LOWvsMOCK_M_df$Gene <- rownames(Day3_LOWvsMOCK_M_df)
Day3_LOWvsMOCK_M_df$Day <- "Day3"
Day3_LOWvsMOCK_M_df$Condition <- "Low"
Day3_LOWvsMOCK_M_df$Sex <- "Male"

Day4_LOWvsMOCK_M <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day4_LOW_M", "Day4_MOCK_M"))
Day4_LOWvsMOCK_M_df <- as.data.frame(Day4_LOWvsMOCK_M)
Day4_LOWvsMOCK_M_df$Gene <- rownames(Day4_LOWvsMOCK_M_df)
Day4_LOWvsMOCK_M_df$Day <- "Day4"
Day4_LOWvsMOCK_M_df$Condition <- "Low"
Day4_LOWvsMOCK_M_df$Sex <- "Male"

Day5_LOWvsMOCK_M <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day5_LOW_M", "Day5_MOCK_M"))
Day5_LOWvsMOCK_M_df <- as.data.frame(Day5_LOWvsMOCK_M)
Day5_LOWvsMOCK_M_df$Gene <- rownames(Day5_LOWvsMOCK_M_df)
Day5_LOWvsMOCK_M_df$Day <- "Day5"
Day5_LOWvsMOCK_M_df$Condition <- "Low"
Day5_LOWvsMOCK_M_df$Sex <- "Male"

# High vs. Mock Female
Day1_HIGHvsMOCK_F <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day1_HIGH_F", "Day1_MOCK_F"))
Day1_HIGHvsMOCK_F_df <- as.data.frame(Day1_HIGHvsMOCK_F)
Day1_HIGHvsMOCK_F_df$Gene <- rownames(Day1_HIGHvsMOCK_F_df)
Day1_HIGHvsMOCK_F_df$Day <- "Day1"
Day1_HIGHvsMOCK_F_df$Condition <- "High"
Day1_HIGHvsMOCK_F_df$Sex <- "Female"

Day2_HIGHvsMOCK_F <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day2_HIGH_F", "Day2_MOCK_F"))
Day2_HIGHvsMOCK_F_df <- as.data.frame(Day2_HIGHvsMOCK_F)
Day2_HIGHvsMOCK_F_df$Gene <- rownames(Day2_HIGHvsMOCK_F_df)
Day2_HIGHvsMOCK_F_df$Day <- "Day2"
Day2_HIGHvsMOCK_F_df$Condition <- "High"
Day2_HIGHvsMOCK_F_df$Sex <- "Female"

Day3_HIGHvsMOCK_F <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day3_HIGH_F", "Day3_MOCK_F"))
Day3_HIGHvsMOCK_F_df <- as.data.frame(Day3_HIGHvsMOCK_F)
Day3_HIGHvsMOCK_F_df$Gene <- rownames(Day3_HIGHvsMOCK_F_df)
Day3_HIGHvsMOCK_F_df$Day <- "Day3"
Day3_HIGHvsMOCK_F_df$Condition <- "High"
Day3_HIGHvsMOCK_F_df$Sex <- "Female"

Day4_HIGHvsMOCK_F <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day4_HIGH_F", "Day4_MOCK_F"))
Day4_HIGHvsMOCK_F_df <- as.data.frame(Day4_HIGHvsMOCK_F)
Day4_HIGHvsMOCK_F_df$Gene <- rownames(Day4_HIGHvsMOCK_F_df)
Day4_HIGHvsMOCK_F_df$Day <- "Day4"
Day4_HIGHvsMOCK_F_df$Condition <- "High"
Day4_HIGHvsMOCK_F_df$Sex <- "Female"

Day5_HIGHvsMOCK_F <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day5_HIGH_F", "Day5_MOCK_F"))
Day5_HIGHvsMOCK_F_df <- as.data.frame(Day5_HIGHvsMOCK_F)
Day5_HIGHvsMOCK_F_df$Gene <- rownames(Day5_HIGHvsMOCK_F_df)
Day5_HIGHvsMOCK_F_df$Day <- "Day5"
Day5_HIGHvsMOCK_F_df$Condition <- "High"
Day5_HIGHvsMOCK_F_df$Sex <- "Female"


# High vs. Mock Male
Day1_HIGHvsMOCK_M <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day1_HIGH_M", "Day1_MOCK_M"))
Day1_HIGHvsMOCK_M_df <- as.data.frame(Day1_HIGHvsMOCK_M)
Day1_HIGHvsMOCK_M_df$Gene <- rownames(Day1_HIGHvsMOCK_M_df)
Day1_HIGHvsMOCK_M_df$Day <- "Day1"
Day1_HIGHvsMOCK_M_df$Condition <- "High"
Day1_HIGHvsMOCK_M_df$Sex <- "Male"

Day2_HIGHvsMOCK_M <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day2_HIGH_M", "Day2_MOCK_M"))
Day2_HIGHvsMOCK_M_df <- as.data.frame(Day2_HIGHvsMOCK_M)
Day2_HIGHvsMOCK_M_df$Gene <- rownames(Day2_HIGHvsMOCK_M_df)
Day2_HIGHvsMOCK_M_df$Day <- "Day2"
Day2_HIGHvsMOCK_M_df$Condition <- "High"
Day2_HIGHvsMOCK_M_df$Sex <- "Male"

Day3_HIGHvsMOCK_M <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day3_HIGH_M", "Day3_MOCK_M"))
Day3_HIGHvsMOCK_M_df <- as.data.frame(Day3_HIGHvsMOCK_M)
Day3_HIGHvsMOCK_M_df$Gene <- rownames(Day3_HIGHvsMOCK_M_df)
Day3_HIGHvsMOCK_M_df$Day <- "Day3"
Day3_HIGHvsMOCK_M_df$Condition <- "High"
Day3_HIGHvsMOCK_M_df$Sex <- "Male"

Day4_HIGHvsMOCK_M <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day4_HIGH_M", "Day4_MOCK_M"))
Day4_HIGHvsMOCK_M_df <- as.data.frame(Day4_HIGHvsMOCK_M)
Day4_HIGHvsMOCK_M_df$Gene <- rownames(Day4_HIGHvsMOCK_M_df)
Day4_HIGHvsMOCK_M_df$Day <- "Day4"
Day4_HIGHvsMOCK_M_df$Condition <- "High"
Day4_HIGHvsMOCK_M_df$Sex <- "Male"

Day5_HIGHvsMOCK_M <- results(DE_2, alpha = 0.05, contrast = c("Collection_Date", "Day5_HIGH_M", "Day5_MOCK_M"))
Day5_HIGHvsMOCK_M_df <- as.data.frame(Day5_HIGHvsMOCK_M)
Day5_HIGHvsMOCK_M_df$Gene <- rownames(Day5_HIGHvsMOCK_M_df)
Day5_HIGHvsMOCK_M_df$Day <- "Day5"
Day5_HIGHvsMOCK_M_df$Condition <- "High"
Day5_HIGHvsMOCK_M_df$Sex <- "Male"

# combine results from all comparisons
DE2_results_LOW_F_df <- rbind(Day1_LOWvsMOCK_F_df, Day2_LOWvsMOCK_F_df, Day3_LOWvsMOCK_F_df, Day4_LOWvsMOCK_F_df, Day5_LOWvsMOCK_F_df)
DE2_results_LOW_M_df <- rbind(Day1_LOWvsMOCK_M_df, Day2_LOWvsMOCK_M_df, Day3_LOWvsMOCK_M_df, Day4_LOWvsMOCK_M_df, Day5_LOWvsMOCK_M_df)

DE2_results_HIGH_F_df <- rbind(Day1_HIGHvsMOCK_F_df, Day2_HIGHvsMOCK_F_df, Day3_HIGHvsMOCK_F_df, Day4_HIGHvsMOCK_F_df, Day5_HIGHvsMOCK_F_df)
DE2_results_HIGH_M_df <- rbind(Day1_HIGHvsMOCK_M_df, Day2_HIGHvsMOCK_M_df, Day3_HIGHvsMOCK_M_df, Day4_HIGHvsMOCK_M_df, Day5_HIGHvsMOCK_M_df)

DE2_results_ALL_df <- rbind(DE2_results_LOW_F_df, DE2_results_LOW_M_df, DE2_results_HIGH_F_df, DE2_results_HIGH_M_df)

DE2_results_F_df <- rbind(Day1_LOWvsMOCK_F_df, Day2_LOWvsMOCK_F_df, Day3_LOWvsMOCK_F_df, Day4_LOWvsMOCK_F_df, Day5_LOWvsMOCK_F_df,
                          Day1_HIGHvsMOCK_F_df, Day2_HIGHvsMOCK_F_df, Day3_HIGHvsMOCK_F_df, Day4_HIGHvsMOCK_F_df, Day5_HIGHvsMOCK_F_df)

DE2_results_M_df <- rbind(Day1_LOWvsMOCK_M_df, Day2_LOWvsMOCK_M_df, Day3_LOWvsMOCK_M_df, Day4_LOWvsMOCK_M_df, Day5_LOWvsMOCK_M_df,
                          Day1_HIGHvsMOCK_M_df, Day2_HIGHvsMOCK_M_df, Day3_HIGHvsMOCK_M_df, Day4_HIGHvsMOCK_M_df, Day5_HIGHvsMOCK_M_df)
```

#### Generate plotting function
```{r}
plot_volcano <- function(DE2_results_ALL_df, day_label, condition_label, colour_label, sex_label) {
  
  # add 'Day' and 'Condition' columns
  DE2_results_ALL_df$Day <- day_label
  DE2_results_ALL_df$Condition <- condition_label
  DE2_results_ALL_df$Sex <- sex_label
  
  # create the volcano plot
  volcano_plot <- ggplot(DE2_results_ALL_df, aes(x = log2FoldChange, y = -log10(pvalue))) +
    geom_point(aes(color = (padj < 0.05  & abs(log2FoldChange) > 1.5)), alpha = 0.6, size = 2) +
    scale_color_manual(values = c("gray", colour_label)) +
    theme_classic() +
    ggtitle(paste(day_label, ":", condition_label, "vs Mock")) +
    xlab(expression(Log[2]~"(Fold Change)")) +
    ylab(expression(-Log[10]~"("*italic(P)*"-value)")) +
    theme(legend.position = "none",
          text = element_text(family = "Times"))
  
  return(volcano_plot)
}
```

#### Genereate combined volcano plot
```{r}
# p-value < 0.05 and Log2FC > |1.5|
DE2_results_ALL_df <- DE2_results_ALL_df %>%
  mutate(sigGroup = ifelse(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1.5,
                           as.character(Condition),
                           "Not significant")) %>%
  mutate(Day = gsub("Day", "Day ", Day))
table(DE2_results_ALL_df$sigGroup)

# plot
combined_volcano_plot_2 <- ggplot(DE2_results_ALL_df, 
                                aes(x = log2FoldChange, y = -log10(pvalue), 
                                    color = sigGroup,
                                    shape = Sex)) +
  geom_point(alpha = 0.2, size = 2) +
  scale_color_manual(
    name = "Condition",
    values = c("High" = "red", "Low" = "blue", "NotSig" = "gray"),
    labels = c("High dose", "Low dose")) +
  scale_shape_manual(
    name = "Sex",
    values = c("Male" = 16, "Female" = 17)) + # 16 = filled circle, 17 = triangle
  facet_wrap(~Day, scales = "free_x") +
  theme_classic() +
  ggtitle("Combined Volcano Plot: Low vs MOCK and High vs MOCK") +
  xlab(expression(Log[2]~"(Fold Change)")) +
  ylab(expression(-Log[10]~"("*italic(P)*"-value)")) +
  theme(text = element_text(family = "Times"),
        strip.text = element_text(size = 11, face = "bold"),
        axis.title.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"))

ggsave("Volcano_Tolerance&Sex_ALL.png", combined_volcano_plot_2, width = 10, height = 7, dpi = 300)

print(combined_volcano_plot_2)
```

#### Generate lethal and tolerant female volcano plot
```{r}
# p-value < 0.05 and Log2FC > |1.5|
DE2_results_F_df <- DE2_results_F_df %>%
  mutate(sigGroup = ifelse(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1.5,
                           as.character(Condition),
                           "Not significant")) %>%
  mutate(Day = gsub("Day", "Day ", Day))
table(DE2_results_F_df$sigGroup)

# plot
combined_volcano_plot_F <- ggplot(DE2_results_F_df, 
                                aes(x = log2FoldChange, y = -log10(pvalue), 
                                    color = sigGroup,
                                    shape = Sex)) +
  geom_point(alpha = 0.2, size = 2) +
  scale_shape_manual(
    name = "Sex",
    values = c("Female" = 17)) + # 16 = filled circle, 17 = triangle
  scale_color_manual(
    name = "Condition",
    values = c("High" = "red", "Low" = "blue", "NotSig" = "gray"),
    labels = c("High dose", "Low dose")) +
  facet_wrap(~Day, scales = "free_x") +
  theme_classic() +
  ggtitle("Volcano Plot: Female - Low vs MOCK and High vs MOCK") +
  xlab(expression(bold(Log[2]~"(Fold Change)"))) +
  ylab(expression(bold(-Log[10]~"("*italic(p)*"-value)"))) +
  theme(text = element_text(family = "Times"),
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 13),
    legend.title = element_blank(),
    legend.text = element_text(size = 13),
    legend.position = "bottom",
    strip.text = element_markdown(size = 14, face = "bold"),  # Enable markdown for color
    panel.border = element_rect(color = "black", fill = NA, size = 0.4),
    axis.line = element_line(size = 0.1, color = "black"),
    strip.background = element_rect(color = "black", fill = "white", size = 0.4)) +
  guides(shape = guide_legend(order = 1),  # Change legend title for color
         color = guide_legend(order = 2))

ggsave("Volcano_Tolerance&Sex_F.png", combined_volcano_plot_F, width = 10, height = 7, dpi = 300)

print(combined_volcano_plot_F)
```

#### Generate lethal and tolerant male volcano plot
```{r}
# p-value < 0.05 and Log2FC > |1.5|
DE2_results_M_df <- DE2_results_M_df %>%
  mutate(sigGroup = ifelse(!is.na(padj) & padj < 0.05 & abs(log2FoldChange) > 1.5,
                           as.character(Condition),
                           "Not significant")) %>%
  mutate(Day = gsub("Day", "Day ", Day))
table(DE2_results_M_df$sigGroup)

# plot
combined_volcano_plot_M <- ggplot(DE2_results_M_df, 
                                aes(x = log2FoldChange, y = -log10(pvalue), 
                                    color = sigGroup,
                                    shape = Sex)) +
  geom_point(alpha = 0.2, size = 2) +
  scale_color_manual(
    name = "Condition",
    values = c("High" = "red", "Low" = "blue", "NotSig" = "gray"),
    labels = c("High dose", "Low dose")) +
  scale_shape_manual(
    name = "Sex",
    values = c("Male" = 16)) + # 16 = filled circle, 17 = triangle
  facet_wrap(~Day, scales = "free_x") +
  theme_classic() +
  ggtitle("Volcano Plot: Male - Low vs MOCK and High vs MOCK") +
  xlab(expression(bold(Log[2]~"(Fold Change)"))) +
  ylab(expression(bold(-Log[10]~"("*italic(p)*"-value)"))) +
  theme(text = element_text(family = "Times"),
    axis.title = element_text(size = 15, face = "bold"),
    axis.text = element_text(size = 13),
    legend.title = element_blank(),
    legend.text = element_text(size = 13),
    legend.position = "bottom",
    strip.text = element_markdown(size = 14, face = "bold"),  # Enable markdown for color
    panel.border = element_rect(color = "black", fill = NA, size = 0.4),
    axis.line = element_line(size = 0.1, color = "black"),
    strip.background = element_rect(color = "black", fill = "white", size = 0.4))

ggsave("Volcano_Tolerance&Sex_M.png", combined_volcano_plot_M, width = 10, height = 7, dpi = 300)

print(combined_volcano_plot_M)
```

## Part II. Generate bar-coded volcano plot

#### Sort DE genes to up- and down-regulated, determine counts
```{r}
# sort DE genes
DE2_results_ALL_df_plot_2 <- DE2_results_ALL_df %>%
  mutate(Regulation = case_when(
    !is.na(padj) & padj < 0.05 & log2FoldChange > 1.5 ~ "Upregulated",
    !is.na(padj) & padj < 0.05 & log2FoldChange < -1.5 ~ "Downregulated",
    TRUE ~ "Not Significant"
  )) %>%
  mutate(Day = gsub("Day", "", Day),
         Group = case_when(
           Condition == "High" & Sex == "Male" ~ "High dose male",
           Condition == "High" & Sex == "Female" ~ "High dose female",
           Condition == "Low" & Sex == "Male" ~ "Low dose male",
           Condition == "Low" & Sex == "Female" ~ "Low dose female"
         ))

# count the number of up- and down-regulated genes per condition and day
bar_data_2 <- DE2_results_ALL_df_plot_2 %>%
  filter(Regulation != "Not Significant") %>%
  group_by(Day, Group, Regulation) %>%
  summarise(Gene_Count = n(), .groups = "drop")
```

### Plot
```{r}
# add labels and colour by condition
bar_data_2 <- bar_data_2 %>%
  mutate(Group_colored = case_when(
    Group == "High dose female" ~ "<span style='color:red;'>High dose female</span>",
    Group == "High dose male" ~ "<span style='color:sienna1;'>High dose male</span>",
    Group == "Low dose female" ~ "<span style='color:blue;'>Low dose female</span>",
    Group == "Low dose male" ~ "<span style='color:cadetblue3;'>Low dose male</span>"
  )) %>%
  mutate(Group_colored = factor(Group_colored, levels = c(
    "<span style='color:red;'>High dose female</span>",
    "<span style='color:sienna1;'>High dose male</span>",
    "<span style='color:blue;'>Low dose female</span>",
    "<span style='color:cadetblue3;'>Low dose male</span>"
  )))

# plot
bar_plot_2 <- ggplot(bar_data_2, aes(x = Day, y = Gene_Count, fill = Regulation)) +
  geom_bar(stat = "identity", position = "stack", width = 0.9) +
  facet_wrap(~ Group_colored, scales = "free_x", nrow=2) +  # Use colored text variable for facets
  scale_fill_manual(values = c("Upregulated" = "orange", "Downregulated" = "purple")) +
  labs(title = "Differentially Expressed Genes Over Time (By Group)",
       x = "Day",
       y = "Gene Count",
       fill = "Regulation") +
  theme_classic() +
  theme(
    text = element_text(family = "Times"),
    axis.title = element_text(size = 20, face = "bold"),
    axis.text = element_text(size = 16),
    legend.title = element_text(size = 17, face = "bold"),
    legend.text = element_text(size = 16),
    legend.position = "bottom",
    strip.text = element_markdown(size = 16, face = "bold"),
    panel.border = element_rect(color = "black", fill = NA, size = 0.4),
    axis.line = element_line(size = 0.1, color = "black"),
    strip.background = element_rect(color = "black", fill = "white", size = 0.4)
  )

ggsave("Bar_Graph_Tolerance&Sex_legend.png", bar_plot_2, width = 7, height = 7, dpi = 300)
print(bar_plot_2)
```
```